Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-03-15 paper/preprint/paper.tex
final rerender of appendix

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be (1) plausible with
respect to the underlying data and (2) actionable with respect to the
user-defined mutability constraints. Much existing research has
therefore focused on developing post-hoc methods to generate
counterfactuals that meet these desiderata. In this work, we instead
hold models directly accountable for the desired end goal:
counterfactual training employs counterfactuals ad-hoc during the
training phase to minimize the divergence between learned
representations and plausible, actionable explanations. We demonstrate
empirically and theoretically that our proposed method facilitates
training models that deliver inherently desirable explanations while
promoting robustness and preserving high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-train-default}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-eval-default}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\caption{\label{tbl-final-params}Final hyperparameters used for the main
results preseneted in Section~\ref{sec-experiments}. Any hyperparameter
not shown here is set to its default value
(Note~\ref{nte-train-default}).}

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-train-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Package Version (Reproducibility)}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-warning-color!10!white, colframe=quarto-callout-warning-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune_lr-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

>>>>

\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be (1) plausible with
respect to the underlying data and (2) actionable with respect to the
user-defined mutability constraints. Much existing research has
therefore focused on developing post-hoc methods to generate
counterfactuals that meet these desiderata. In this work, we instead
hold models directly accountable for the desired end goal:
counterfactual training employs counterfactuals ad-hoc during the
training phase to minimize the divergence between learned
representations and plausible, actionable explanations. We demonstrate
empirically and theoretically that our proposed method facilitates
training models that deliver inherently desirable explanations while
promoting robustness and preserving high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-train-default}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-eval-default}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\caption{\label{tbl-final-params}Final hyperparameters used for the main
results presented in Section~\ref{sec-experiments}. Any hyperparameter
not shown here is set to its default value
(Note~\ref{nte-train-default}).}

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-warning-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-warning-color!10!white, title={Package Version (Reproducibility)}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
